Enforced Invariant Confluence

### 2018-04-22 doc/paper/sections/merge_reduction.tex
Formatting tweaks.

\section{Merge Reduction}\seclabel{MergeReduction}
In \secref{InvariantClosure}, we discussed how invariant-closure is
fundamentally a property of reachability and that invariant-closure is
sufficient but not necessary for invariant-confluence because it fails to
incorporate any notion of reachability. Using this intuition, we established
\thmref{IclosureEquivalentIconfluence} and then exploited the theorem in
\algoref{InteractiveDecisionProcedure}. In this section, we again take
advantage of this intuition to develop a new sufficient condition for
invariant-confluence that can be checked without user interaction and that
covers some cases not covered by invariant-closure.

An expression $e = t_1(t_2(\ldots(t_n(s))\ldots))$ is \defword{merge-free} if
does not contain any merges (i.e.\ it is generated by the grammar $e ::= s \mid
t(e)$). An object $O = (S, \join)$ is \defword{merge-reducible} with respect to
a start state $s_0 \in S$, a set of transactions $T$, and an invariant $I$,
abbreviated \defword{\sTImergereducible}, if for every pair of $e_1$ and $e_2$
of merge-free $\sTIreachable{}$ expressions, there exists some merge-free
$\sTIreachable{}$ expression $e_3$ that evaluates to the same state as $e_1
\join e_2$. Intuitively, if $O$ is merge-reducible, we can replace $e_1 \join
e_2$ (which has one merge) with $e_3$ (which has no merges) to obtain an
equivalent expression with fewer merges.

@@

The proof of \thmref{ReducibilityImpliesIconfluence} is a straightforward
induction. We begin with an $\sTIreachable{}$ expression $e$ and repeatedly
replace any subexpression that merges two merge-free subexpressions with an
equivalent merge-free reachable subexpression (which we can do because $O$ is
merge-reducible). We repeat this process until $e$ has been completely replaced
with an equivalent merge-free reachable expression $e'$. Because $I(s_0)$ and
because our system model only executes transactions that preserve the
invariant, $e'$ (and hence $e$) is guaranteed to satisfy the invariant. Thus,
all reachable states satisfy the invariant, so $O$ is invariant-confluent.

Merge-reducibility is a sufficient condition for invariant-confluence, but
unlike with invariant-closure, it is not straightforward to automatically
determine if an object is merge-reducible. In \thmref{LatticeProperty}, we
outline a sufficient condition for merge-reducibility that is straightforward
to determine automatically.

@@

\thmref{IclosureImpliesIconfluence} states that invariant-closure is a
sufficient condition for invariant-confluence, and \thmref{LatticeProperty}
states that criteria (1) -- (4) are sufficient conditions for
invariant-confluence. How do these sufficient conditions relate to one another?
Clearly, not all invariant-closed objects are semilattices, so
invariant-closure does not imply criteria (1) -- (4). Conversely, there are some
objects which satisfy criteria (1) -- (4) which are not invariant-closed. Here's
one example.

@@

Invariant-closure is not necessary for invariant-confluence because it fails to
incorporate any notion of reachability. Criteria (1) -- (4) are also
unnecessary, but they can be used to prove that some non-invariant-closed
objects are invariant-confluent because the criteria \emph{do} incorporate
notions of reachability. In particular, criterion (3) is a slight variant of
invariant-closure; it also states that invariant-satisfying states should be
closed under merge. The fundamental difference is that criterion (3) restricts
its attention to the merge of two states which are \emph{reachable} from a
common ancestor state.

In \exampleref{TwoSets}, we saw this fundamental difference rear its head. $O$
is not $I$-closed because the union of an odd-only set with an even-only set is
a set with both odd and even integers. However, if we begin in an
invariant-satisfying state, we cannot reach both an odd-only and even-only set.
Criterion (3) is able to recognize this fact and conclude that $O$ is
invariant-confluent despite it not being invariant-closed.

>>>>

\section{Merge Reduction}\seclabel{MergeReduction}
In \secref{InvariantClosure}, we discussed how \invariantclosure{} is
fundamentally a property of reachability and that \invariantclosure{} is
sufficient but not necessary for \invariantconfluence{} because it fails to
incorporate any notion of reachability. Using this intuition, we established
\thmref{IclosureEquivalentIconfluence} and then exploited the theorem in
\algoref{InteractiveDecisionProcedure}. In this section, we again take
advantage of this intuition to develop a new sufficient condition for
\invariantconfluence{} that can be checked without user interaction and that
covers some cases not covered by \invariantclosure{}.

An expression $e = t_1(t_2(\ldots(t_n(s))\ldots))$ is \defword{merge-free} if
does not contain any merges (i.e.\ it is generated by the grammar $e ::= s \mid
t(e)$). An object $O = (S, \join)$ is \defword{merge-reducible} with respect to
a start state $s_0 \in S$, a set of transactions $T$, and an invariant $I$,
abbreviated \defword{\sTImergereducible}, if for every pair of $e_1$ and $e_2$
of merge-free \sTIreachable{} expressions, there exists some merge-free
\sTIreachable{} expression $e_3$ that evaluates to the same state as $e_1
\join e_2$. Intuitively, if $O$ is merge-reducible, we can replace $e_1 \join
e_2$ (which has one merge) with $e_3$ (which has no merges) to obtain an
equivalent expression with fewer merges.

@@

The proof of \thmref{ReducibilityImpliesIconfluence} is a straightforward
induction. We begin with an \sTIreachable{} expression $e$ and repeatedly
replace any subexpression that merges two merge-free subexpressions with an
equivalent merge-free reachable subexpression (which we can do because $O$ is
merge-reducible). We repeat this process until $e$ has been completely replaced
with an equivalent merge-free reachable expression $e'$. Because $I(s_0)$ and
because our system model only executes transactions that preserve the
invariant, $e'$ (and hence $e$) is guaranteed to satisfy the invariant. Thus,
all reachable states satisfy the invariant, so $O$ is \invariantconfluent{}.

Merge-reducibility is a sufficient condition for \invariantconfluence{}, but
unlike with \invariantclosure{}, it is not straightforward to automatically
determine if an object is merge-reducible. In \thmref{LatticeProperty}, we
outline a sufficient condition for merge-reducibility that is straightforward
to determine automatically.

@@

\thmref{IclosureImpliesIconfluence} states that \invariantclosure{} is a
sufficient condition for \invariantconfluence{}, and \thmref{LatticeProperty}
states that criteria (1) -- (4) are sufficient conditions for
\invariantconfluence{}. How do these sufficient conditions relate to one another?
Clearly, not all \invariantclosed{} objects are semilattices, so
\invariantclosure{} does not imply criteria (1) -- (4). Conversely, there are some
objects which satisfy criteria (1) -- (4) which are not \invariantclosed{}. Here's
one example.

@@

\Invariantclosure{} is not necessary for \invariantconfluence{} because it fails to
incorporate any notion of reachability. Criteria (1) -- (4) are also
unnecessary, but they can be used to prove that some non-\invariantclosed{}
objects are \invariantconfluent{} because the criteria \emph{do} incorporate
notions of reachability. In particular, criterion (3) is a slight variant of
\invariantclosure{}; it also states that invariant satisfying states should be
closed under merge. The fundamental difference is that criterion (3) restricts
its attention to the merge of two states which are \emph{reachable} from a
common ancestor state.

In \exampleref{TwoSets}, we saw this fundamental difference rear its head. $O$
is not $I$-closed because the union of an odd-only set with an even-only set is
a set with both odd and even integers. However, if we begin in an
invariant satisfying state, we cannot reach both an odd-only and even-only set.
Criterion (3) is able to recognize this fact and conclude that $O$ is
\invariantconfluent{} despite it not being \invariantclosed{}.

### 2018-04-22 doc/paper/sections/related_work.tex
Formatting tweaks.

\newcommand{\sieve}{\textsc{SIEVE}}
\textbf{RedBlue Consistency and \sieve.}
RedBlue consistency is a consistency model that sits between causal consistency
and linearizability~\cite{li2012making}.
%
With RedBlue consistency, every operation is manually labelled as either red or
blue. All operations are executed with causal consistency, but with the added
restrictions that red operations are executed in a single total order embedded
within the causal ordering.
%
Invariant safety is a sufficient condition for RedBlue consistent objects to be
invariant-confluent. Invariant safety is an analog of invariant-closure for an
operation-based system model (as opposed-to the state based system model that
we use). In~\cite{li2014automating}, Li et al.\ develop techniques to identify
invariant safe operations. Like invariant-closure, invariant safety is a
sufficient but not necessary condition for invariant-confluence. Li et al.\
develop sophisticated techniques for deciding invariant safety that involve
calculating weakest preconditions. These techniques are complementary to our
work and can be used to improve the invariant-closure subroutine used by our
decision procedures.

\textbf{The Homeostasis Protocol.}
The homeostasis protocol~\cite{roy2015homeostasis} uses program analysis to
avoid unnecessary coordination between servers in a \emph{sharded} database
(whereas invariant-confluence targets \emph{replicated} databases). The
protocol generates global invariants, called global treaties, and servers can
operate without coordination so long as they preserve the treaty. When the
treaty is violated, the nodes perform a round of coordination and establish a
new treaty. The protocol guarantees that transactions are executed with
observational equivalence with respect to some serial execution of the
transactions. This means that intermediate states may be inconsistent, but
externally observable side effects and the final database state are consistent.
The observational equivalence guaranteed by the homeostasis is stronger than
the guarantees of invariant-confluence. As a result, there are invariants and
workloads that the homeostasis protocol would execute with more coordination
than a segmented invariant-confluent execution. Still, the homeostasis
protocol's mechanism of establishing invariants and operating without
coordination so long as the invariants are maintained is very similar to our
design of segmented invariant-confluence.

\textbf{Explicit Consistency.}
Explicit consistency~\cite{balegas2015towards} is a consistency model that
combines invariant-confluence and causal consistency, similar to RedBlue
consistency with invariant safety. To determine if a workload is amenable to
explicitly consistent replication, Balegas et al.\ determine if all pairs of
transactions can be concurrently executed on the same start state without
violating the invariant~\cite{balegas2015towards}.  This is a sufficient
condition for explicit consistency similar to criterion (3) in
\thmref{LatticeProperty}. Balegas et al.\ also describe a variety of
techniques---like conflict resolution, locking, and escrow
transactions~\cite{o1986escrow}---that can be used to replicate workloads that
do not meet their sufficient conditions. As discussed in \secref{Evaluation},
simple forms of these techniques can be modelled with segmented
invariant-confluence.

\textbf{Token-Based Invariant-Confluence.}
In~\cite{gotsman2016cause}, Gotsman et al.\ discuss a hybrid token-based
consistency model that generalizes a family of consistency models including
causal consistency, sequential consistency, and RedBlue consistency. An
application designer defines a set of tokens and specifies which pairs of
tokens conflict, and transactions acquire some subset of the tokens when they
execute. This allows the application designer to specify which transactions
conflict with each other.
%
% With no tokens, the model is equivalent to causal consistency. With all
% transactions acquiring a single self-conflicting token, the model is equivalent
% to sequential consistency~\cite{lamport1979make}. With some transactions
% acquiring a single self-conflicting token and some transactions acquiring no
% tokens, the model is equivalent to RedBlue consistency.
%
Gotsman et al.\ develop sufficient conditions to determine whether a given
token scheme is sufficient to guarantee that a global invariant is never
broken. This model of consistency is very similar to segmented
invariant-confluence. The token-based approach allows users to specify certain
conflicts that are not possible with segmented invariant-confluence. A
segmentation only allows transactions within a segment to acquire a single
self-conflicting lock. However, segmented invariant-confluence also introduces
the notion of invariant segmentation, which cannot be emulated with the token
based approach. For example, it is difficult to emulate escrow transactions
with the token-based approach.

\textbf{CRDTs.}
CRDTs~\cite{shapiro2011conflict, shapiro2011comprehensive} are distributed
semilattices with inflationary update methods. Due to their algebraic
properties, CRDTs can be replicated with strong eventual consistency without
the need for any coordination. Our definition of distributed objects and our
invariant-confluence system model are inspired directly by the corresponding
definitions and system models in~\cite{shapiro2011conflict}. Note that an
invariant-confluent object does not necessarily have to be eventually
consistent, so it is natural (though not necessary) to use CRDTs as
invariant-confluent distributed objects, achieving a combination of eventual
consistency and invariant-preservation. Our criteria in
\thmref{LatticeProperty} also borrow ideas from CRDTs, exploiting the nice
algebraic properties of semilattices.

\textbf{CALM Theorem.}
Bloom~\cite{alvaro2010boom, alvaro2011consistency, conway2012logic} and its
formalism, Dedalus~\cite{alvaro2011dedalus, alvaro2013declarative}, are
declarative Datalog-based programming languages that are designed to program
distributed systems. The accompanying CALM
theorem~\cite{hellerstein2010declarative, ameloot2013relational} states that if
a program can be written in the monotone fragment of these languages, then
there exists a coordination-free implementation of the program.
% Moreover, monotone programs are invariant to message reordering, whereas
% non-monotone programs may experience anomalies if messages are reordered in the
% network.
The CALM theorem provides guarantees about the outputs of a program, but it
does not guarantee that invariants are maintained throughout the duration of an
execution. Moreover, Bloom and Dedalus are general purpose programming
languages that can be used to implement a variety of distributed systems (e.g.
a sharded database) which are outside of the scope of invariant-confluence.

\textbf{Program Analysis in Database Systems.}
Using program analysis to improve the performance of database systems is not
new. For example, it has been used to improve the performance of
database-backed web applications~\cite{cheung2014using, wu2016transaction,
ramachandra2012program} and used to improve the performance of optimistic
concurrency control on multi-core machines~\cite{wu2016transaction}. Our work
on invariant-confluence continues the theme of using program analysis to reap
the performance benefits gained from understanding program semantics.

% \textbf{Proof Assistants.}
% Our interactive decision procedure bears resemblance to interactive proof
% assistants like Coq~\cite{coq2017}, Isabelle~\cite{nipkow2002isabelle},
% Agda~\cite{norell2008dependently}, and Idris~\cite{brady2013idris}. Proof
% assistants guide users through the process of interactively provided a proof
% that witnesses the validity of a formula. Our interactive decision procedure
% guides users through the process of interactively generating a set of reachable
% and unreachable points that witness the invariant-confluence of an object. An
% interesting avenue for future work is to integrate an proof assistant into our
% interactive decision procedure. For example, a proof assistant could verify
% that a state labelled unreachable by a user is actually unreachable.

>>>>

\newcommand{\sieve}{\textsc{SIEVE}}
\textbf{RedBlue Consistency and \sieve.}
RedBlue consistency is a consistency model that sits between causal consistency
and linearizability~\cite{li2012making}.
%
With RedBlue consistency, every operation is manually labelled as either red or
blue. All operations are executed with causal consistency, but with the added
restrictions that red operations are executed in a single total order embedded
within the causal ordering.
%
Invariant safety is a sufficient condition for RedBlue consistent objects to be
\invariantconfluent{}. Invariant safety is an analog of \invariantclosure{} for an
operation-based system model (as opposed-to the state based system model that
we use). In~\cite{li2014automating}, Li et al.\ develop techniques to identify
invariant safe operations. Like \invariantclosure{}, invariant safety is a
sufficient but not necessary condition for \invariantconfluence{}. Li et al.\
develop sophisticated techniques for deciding invariant safety that involve
calculating weakest preconditions. These techniques are complementary to our
work and can be used to improve the \invariantclosure{} subroutine used by our
decision procedures.

\textbf{The Homeostasis Protocol.}
The homeostasis protocol~\cite{roy2015homeostasis} uses program analysis to
avoid unnecessary coordination between servers in a \emph{sharded} database
(whereas \invariantconfluence{} targets \emph{replicated} databases). The
protocol generates global invariants, called global treaties, and servers can
operate without coordination so long as they preserve the treaty. When the
treaty is violated, the nodes perform a round of coordination and establish a
new treaty. The protocol guarantees that transactions are executed with
observational equivalence with respect to some serial execution of the
transactions. This means that intermediate states may be inconsistent, but
externally observable side effects and the final database state are consistent.
The observational equivalence guaranteed by the homeostasis is stronger than
the guarantees of \invariantconfluence{}. As a result, there are invariants and
workloads that the homeostasis protocol would execute with more coordination
than a segmented \invariantconfluent{} execution. Still, the homeostasis
protocol's mechanism of establishing invariants and operating without
coordination so long as the invariants are maintained is very similar to our
design of segmented \invariantconfluence{}.

\textbf{Explicit Consistency.}
Explicit consistency~\cite{balegas2015towards} is a consistency model that
combines \invariantconfluence{} and causal consistency, similar to RedBlue
consistency with invariant safety. To determine if a workload is amenable to
explicitly consistent replication, Balegas et al.\ determine if all pairs of
transactions can be concurrently executed on the same start state without
violating the invariant~\cite{balegas2015towards}.  This is a sufficient
condition for explicit consistency similar to criterion (3) in
\thmref{LatticeProperty}. Balegas et al.\ also describe a variety of
techniques---like conflict resolution, locking, and escrow
transactions~\cite{o1986escrow}---that can be used to replicate workloads that
do not meet their sufficient conditions. As discussed in \secref{Evaluation},
simple forms of these techniques can be modelled with segmented
\invariantconfluence{}.

\textbf{Token Based Invariant Confluence.}
In~\cite{gotsman2016cause}, Gotsman et al.\ discuss a hybrid token-based
consistency model that generalizes a family of consistency models including
causal consistency, sequential consistency, and RedBlue consistency. An
application designer defines a set of tokens and specifies which pairs of
tokens conflict, and transactions acquire some subset of the tokens when they
execute. This allows the application designer to specify which transactions
conflict with each other.
%
% With no tokens, the model is equivalent to causal consistency. With all
% transactions acquiring a single self-conflicting token, the model is equivalent
% to sequential consistency~\cite{lamport1979make}. With some transactions
% acquiring a single self-conflicting token and some transactions acquiring no
% tokens, the model is equivalent to RedBlue consistency.
%
Gotsman et al.\ develop sufficient conditions to determine whether a given
token scheme is sufficient to guarantee that a global invariant is never
broken. This model of consistency is very similar to segmented
\invariantconfluence{}. The token-based approach allows users to specify certain
conflicts that are not possible with segmented \invariantconfluence{}. A
segmentation only allows transactions within a segment to acquire a single
self-conflicting lock. However, segmented \invariantconfluence{} also introduces
the notion of invariant segmentation, which cannot be emulated with the token
based approach. For example, it is difficult to emulate escrow transactions
with the token-based approach.

\textbf{CRDTs.}
CRDTs~\cite{shapiro2011conflict, shapiro2011comprehensive} are distributed
semilattices with inflationary update methods. Due to their algebraic
properties, CRDTs can be replicated with strong eventual consistency without
the need for any coordination. Our definition of distributed objects and our
\invariantconfluence{} system model are inspired directly by the corresponding
definitions and system models in~\cite{shapiro2011conflict}. Note that an
\invariantconfluent{} object does not necessarily have to be eventually
consistent, so it is natural (though not necessary) to use CRDTs as
\invariantconfluent{} distributed objects, achieving a combination of eventual
consistency and invariant preservation. Our criteria in
\thmref{LatticeProperty} also borrow ideas from CRDTs, exploiting the nice
algebraic properties of semilattices.

\textbf{CALM Theorem.}
Bloom~\cite{alvaro2010boom, alvaro2011consistency, conway2012logic} and its
formalism, Dedalus~\cite{alvaro2011dedalus, alvaro2013declarative}, are
declarative Datalog-based programming languages that are designed to program
distributed systems. The accompanying CALM
theorem~\cite{hellerstein2010declarative, ameloot2013relational} states that if
a program can be written in the monotone fragment of these languages, then
there exists a coordination-free implementation of the program.
% Moreover, monotone programs are invariant to message reordering, whereas
% non-monotone programs may experience anomalies if messages are reordered in the
% network.
The CALM theorem provides guarantees about the outputs of a program, but it
does not guarantee that invariants are maintained throughout the duration of an
execution. Moreover, Bloom and Dedalus are general purpose programming
languages that can be used to implement a variety of distributed systems (e.g.
a sharded database) which are outside of the scope of \invariantconfluence{}.

\textbf{Program Analysis in Database Systems.}
Using program analysis to improve the performance of database systems is not
new. For example, it has been used to improve the performance of
database-backed web applications~\cite{cheung2014using, wu2016transaction,
ramachandra2012program} and used to improve the performance of optimistic
concurrency control on multi-core machines~\cite{wu2016transaction}. Our work
on \invariantconfluence{} continues the theme of using program analysis to reap
the performance benefits gained from understanding program semantics.

% \textbf{Proof Assistants.}
% Our interactive decision procedure bears resemblance to interactive proof
% assistants like Coq~\cite{coq2017}, Isabelle~\cite{nipkow2002isabelle},
% Agda~\cite{norell2008dependently}, and Idris~\cite{brady2013idris}. Proof
% assistants guide users through the process of interactively provided a proof
% that witnesses the validity of a formula. Our interactive decision procedure
% guides users through the process of interactively generating a set of reachable
% and unreachable points that witness the \invariantconfluence{} of an object. An
% interesting avenue for future work is to integrate an proof assistant into our
% interactive decision procedure. For example, a proof assistant could verify
% that a state labelled unreachable by a user is actually unreachable.
